---
title: "GETTING INTO REAL ESTATE BUSINESS IN THE US"
author: "Maurice otieno"
date: "2025-05-16"
output: html_document
---

This project is geared towards extracting and analyzing data on housing prices in the United States, with the aim of guiding individuals interested in real estate investment. An in-depth understanding of property pricing dynamics is essential for making data-driven decisions that lead to profitable investments.
This study examines three key assumptions: (1) larger lot sizes and higher counts of bedrooms and bathrooms are positively associated with property prices; (2) homes featuring waterfront views tend to command premium sale values; and (3) properties situated in more developed cities exhibit higher average prices. These assumptions will be tested using statistical methods applied to cleaned and structured data.

To support this analysis, a publicly available dataset from [Kaggle](https://www.kaggle.com/datasets/fratzcan/usa-house-prices) is used as the basis for this exploration.
The dataset was downloaded to a local machine and uploaded to Google Colab for analysis. The ZIP file for the downloaded dataset can be found at this [GitHub repository](https://github.com/otienomaurice1/Getting_Into_Business.git) for inclusion as well.

# 1. DATA UNDERSTANDING

The time range of the dataset spans from **2014 to 2015**, as indicated by the `date` column, which represents the sale date for each property. This provides a snapshot of the housing market during that period.
The data was most likely compiled from **public property records**, listing services, or real estate transaction databases. While the exact collection method is not explicitly documented. This dataset  contains the following  information on residential properties in the United States.

```{r house-prices-from-zip, echo=FALSE, message=FALSE, warning=FALSE}
# Install required packages silently
if (!require(readr)) install.packages("readr")

# Set download paths
zip_url <- "https://github.com/otienomaurice1/Getting_Into_Business/raw/main/Business.zip"
zip_dest <- "Business.zip"
extract_folder <- "usa_house_prices"

# Download and extract ZIP
download.file(zip_url, destfile = zip_dest, mode = "wb")
if (!dir.exists(extract_folder)) dir.create(extract_folder)
unzip(zip_dest, exdir = extract_folder)

# Load CSV
csv_files <- list.files(extract_folder, pattern = "\\.csv$", full.names = TRUE)
if (length(csv_files) == 0) stop("No CSV file found in the extracted folder.")
df <- readr::read_csv(csv_files[1])
```



- **Date**: The date when the property was sold. This feature helps in understanding the temporal trends in property prices.

- **Price**: The sale price of the property in USD. This is the target variable we aim to predict.

- **Bedrooms**: The number of bedrooms in the property. Generally, properties with more bedrooms tend to have higher prices.

- **Bathrooms**: The number of bathrooms in the property. Similar to bedrooms, more bathrooms can increase a property’s value.

- **Sqft Living**: The size of the living area in square feet. Larger living areas are typically associated with higher property values.

- **Sqft Lot**: The size of the lot in square feet. Larger lots may increase a property’s desirability and value.

- **Floors**: The number of floors in the property. Properties with multiple floors may offer more living space and appeal.

- **Waterfront**: A binary indicator (1 if the property has a waterfront view, 0 otherwise). Properties with waterfront views are often valued higher.

- **View**: An index from 0 to 4 indicating the quality of the property’s view. Better views are likely to enhance a property’s value.

- **Condition**: An index from 1 to 5 rating the condition of the property. Properties in better condition are typically worth more.

- **Sqft Above**: The square footage of the property above the basement. This can help isolate the value contribution of above-ground space.

- **Sqft Basement**: The square footage of the basement. Basements may add value depending on their usability.

- **Yr Built**: The year the property was built. Older properties may have historical value, while newer ones may offer modern amenities.

- **Yr Renovated**: The year the property was last renovated. Recent renovations can increase a property’s appeal and value.

- **Street**: The street address of the property. This feature can be used to analyze location-specific price trends.

- **City**: The city where the property is located. Different cities have distinct market dynamics.

- **Statezip**: The state and zip code of the property. This feature provides regional context for the property.

- **Country**: The country where the property is located. While this dataset focuses on properties in Australia, this feature is included for completeness.

```{r data_categorization_table, echo=FALSE, message=FALSE, warning=FALSE}
# Install required packages
if (!require(dplyr)) install.packages("dplyr")
if (!require(knitr)) install.packages("knitr")

library(dplyr)
library(knitr)

# Convert date column properly to Date object
if ("date" %in% names(df)) {
  df$date <- suppressWarnings(as.Date(df$date, format = "%Y-%m-%d %H:%M:%S"))
}

# Define dataset metadata
attributes <- c(
  "date", "price", "bedrooms", "bathrooms", "sqft_lot",
  "floors", "waterfront", "view", "condition",
  "sqft_above", "sqft_basement", "yr_built", "yr_renovated",
  "street", "city", "statezip", "country"
)

types <- c(
  "Categorical", "Numerical", "Numerical", "Numerical", "Numerical",
  "Numerical", "Categorical", "Ordinal", "Ordinal",
  "Numerical", "Numerical", "Numerical", "Numerical",
  "Categorical", "Categorical", "Categorical", "Categorical"
)

levels <- c(
  "Interval", "Ratio", "Ratio", "Ratio", "Ratio",
  "Ratio", "Nominal", "Ordinal", "Ordinal",
  "Ratio", "Ratio", "Interval", "Interval",
  "Nominal", "Nominal", "Nominal", "Nominal"
)

# Attributes where 0 should NOT be considered missing
zero_allowed <- c("waterfront", "view", "condition")

# Count missing and valid entries using custom logic
counts <- sapply(attributes, function(attr) {
  values <- df[[attr]]
  is_missing <- is.na(values) |
                values == "" |
                (attr %in% attributes & !(attr %in% zero_allowed) &
                 suppressWarnings(as.numeric(values) == 0))
  sum(!is_missing)
})

missing <- sapply(attributes, function(attr) {
  values <- df[[attr]]
  is_missing <- is.na(values) |
                values == "" |
                (attr %in% attributes & !(attr %in% zero_allowed) &
                 suppressWarnings(as.numeric(values) == 0))
  sum(is_missing)
})

# Build final categorization table
feature_table <- data.frame(
  Attribute = attributes,
  Type = types,
  Level_of_Measurement = levels,
  Count = counts,
  Missing = missing,
  stringsAsFactors = FALSE
)

# Display GitHub-compatible table
knitr::kable(feature_table, caption = "Attribute Categorization with Custom Missing Value Counts", row.names = FALSE)
```


Based on the data, most attributes have complete values, including date, sqft_lot, floors, and all categorical fields such as waterfront, view, condition, street, city, statezip, and country. The price column has 49 missing values, which is notable since it's the target variable and essential for modeling. bedrooms and bathrooms have only 2 missing entries each, indicating high completeness. However, substantial missingness is present in sqft_basement and yr_renovated, with 2,476 and 2,462 missing entries respectively, suggesting these features may be  frequently unrecorded in the dataset.

To prepare the dataset for analysis, we removed records with missing or zero values in key numerical fields such as `price`, as these entries either lacked meaningful information. For categorical variables like `city` and `street`, empty strings were treated as missing.  `yr_renovated` was left as is because it didn't affect our analysis. This cleaning process ensured that statistical summaries and visualizations were based on accurate, complete data, thereby improving the reliability of insights drawn from the dataset.

# 2. EXPLORATORY DATA ANALYSIS

```{r ratio_summary_tests, echo=FALSE, message=FALSE, warning=FALSE}
# Load required libraries
if (!require(dplyr)) install.packages("dplyr")
if (!require(knitr)) install.packages("knitr")

library(dplyr)
library(knitr)

# Define ratio-level attributes
ratio_attributes <- c(
  "price", "bedrooms", "bathrooms", "sqft_lot",
  "floors", "sqft_above", "sqft_basement", "yr_built", "yr_renovated"
)

# Filter data to remove rows with NA or 0 values in ratio columns
clean_ratio_df <- df %>%
  select(all_of(ratio_attributes)) %>%
  filter(if_all(everything(), ~ !is.na(.) & . != 0))

# Initialize summary table
summary_table <- data.frame()

# Generate summary stats per attribute
for (col in ratio_attributes) {
  col_data <- clean_ratio_df[[col]]
  min_val <- min(col_data)
  max_val <- max(col_data)
  summary_table <- rbind(summary_table, data.frame(
    Attribute = col,
    Count = length(col_data),
    Mean = mean(col_data),
    Median = median(col_data),
    Min = min_val,
    Max = max_val,
    Range = max_val - min_val,
    row.names = NULL
  ))
}

# Render table
knitr::kable(summary_table, digits = 2, caption = "Summary Statistics for Ratio-Level Attributes (Including Range)")
```

The price attribute shows a wide range, indicating substantial variation in property values, which is expected in a housing market. Both bedrooms and bathrooms exhibit relatively low variability, with most properties having a moderate number of rooms. sqft_lot shows high dispersion between the minimum and maximum, suggesting that while most homes sit on average-sized lots, a few properties occupy very large plots. The attributes floors, sqft_above, and sqft_basement reflect architectural diversity, with some homes having multiple stories and finished basements, while others are more compact. yr_built spans a long historical range, highlighting the mix of both older and newer properties. In contrast, yr_renovated has a narrower distribution and lower counts, indicating that fewer homes have undergone recent updates.

To evaluate the validity of the proposed assumptions, we will generate a series of visualizations that explore the relationships between property price and various features such as lot size, number of rooms, renovation status, waterfront presence, and city location. 

We want to investigate whether larger lot sizes are associated with higher property prices.
Since both sqft_lot and price are continuous numeric variables, the most appropriate plot is a scatter plot with a regression line to visualize any linear trend.


```{r plot_lot_size_price, echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)

# Filter out invalid or missing values
lot_df <- df %>%
  filter(price > 0, sqft_lot > 0)

# Scatter plot with regression line
ggplot(lot_df, aes(x = sqft_lot, y = price)) +
  geom_point(alpha = 0.3, color = "darkgreen") +
  geom_smooth(method = "lm", color = "blue", se = FALSE) +
  theme_minimal() +
  labs(
    title = "Relationship Between Lot Size and Price",
    x = "Lot Size (sqft)",
    y = "Price (USD)"
  )
```


The scatter plot shows a positive but weak relationship between lot size and price. While properties with very large lots tend to have higher prices in some cases, the majority of points are clustered around moderate lot sizes and prices, suggesting that lot size alone is not a strong predictor of property value. The regression line indicates a slight upward trend, but the wide spread of values shows that other variables likely contribute more significantly to pricing.


We also aim to determine whether properties with more bedrooms generally sell for higher prices.
Because bedrooms is a discrete numeric variable and price is continuous, a boxplot is the best choice. This will let us visualize how the distribution of prices changes with the number of bedrooms, and whether there's a clear price premium as bedroom count increases.


```{r plot_bedrooms_price, echo=FALSE, message=FALSE, warning=FALSE}
# Filter dataset
bedroom_df <- df %>%
  filter(price > 0, bedrooms > 0)

# Boxplot of Price by Number of Bedrooms
ggplot(bedroom_df, aes(x = as.factor(bedrooms), y = price, fill = as.factor(bedrooms))) +
  geom_boxplot() +
  theme_minimal() +
  labs(
    title = "Price Distribution by Number of Bedrooms",
    x = "Number of Bedrooms",
    y = "Price (USD)"
  ) +
  theme(legend.position = "none")
```


The plot shows that median prices generally increase from 1 to 5 bedrooms, indicating a positive association between bedroom count and property price. However, beyond 5 bedrooms, the trend becomes less consistent. Properties with 6 or more bedrooms do not show significantly higher median prices and tend to have greater variability, likely due to niche market segments or custom-built homes. Overall, 3-bedroom and 4-bedroom homes are the most common, and they occupy the mid-range in both price and spread. This supports the idea that bedroom count influences price, but only up to a practical limit, after which the relationship weakens.

We want to determine whether homes with waterfront views (coded as 1) sell for higher prices compared to those without (0). Since this is a binary categorical variable, a boxplot comparing price distributions between the two groups is ideal.
```{r plot_waterfront_price, echo=FALSE, message=FALSE, warning=FALSE}
# Prepare data
waterfront_df <- df %>%
  filter(price > 0, !is.na(waterfront)) %>%
  mutate(waterfront = ifelse(waterfront == 1, "Waterfront", "No Waterfront"))

# Boxplot of Price by Waterfront Presence
ggplot(waterfront_df, aes(x = waterfront, y = price, fill = waterfront)) +
  geom_boxplot() +
  theme_minimal() +
  scale_fill_manual(values = c("No Waterfront" = "#FFCC99", "Waterfront" = "#99CCFF")) +
  labs(
    title = "Price Distribution by Waterfront Presence",
    x = "Waterfront View",
    y = "Price (USD)"
  ) +
  theme(legend.position = "none")
```


The boxplot reveals a clear price premium for properties with waterfront views. Homes classified as "Waterfront" exhibit significantly higher median prices and overall higher price ranges compared to non-waterfront properties. Although there are far fewer waterfront listings, their elevated price levels strongly support the hypothesis that a waterfront view substantially increases property value.

We want to see whether properties in certain cities consistently have higher or lower prices. We'll group cities by their median property price, then: Show the top 10 cities with the highest medians and Show the bottom 10 cities with the lowest medians


```{r plot_city_price_facets, echo=FALSE, message=FALSE, warning=FALSE}
# Prepare and filter data
city_df <- df %>%
  filter(price > 0, !is.na(city)) %>%
  group_by(city) %>%
  mutate(median_price = median(price)) %>%
  ungroup()

# Get top 10 and bottom 10 cities by median price
top_cities <- city_df %>%
  distinct(city, median_price) %>%
  arrange(desc(median_price)) %>%
  slice_head(n = 10) %>%
  pull(city)

bottom_cities <- city_df %>%
  distinct(city, median_price) %>%
  arrange(median_price) %>%
  slice_head(n = 10) %>%
  pull(city)

# Label groups
city_df <- city_df %>%
  mutate(
    group = case_when(
      city %in% top_cities ~ "Top 10 Cities",
      city %in% bottom_cities ~ "Bottom 10 Cities",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(group))

# Plot
ggplot(city_df, aes(x = reorder(city, price, FUN = median), y = price, fill = city)) +
  geom_boxplot(outlier.size = 0.5) +
  facet_wrap(~group, scales = "free_y") +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Price Distribution in Top and Bottom 10 Cities by Median Price",
    x = "City",
    y = "Price (USD)"
  ) +
  theme(legend.position = "none")
```



The top 10 cities with the highest median property prices include Medina, Yarrow Point, Clyde Hill, Mercer Island, and Bellevue — all affluent suburbs located in or near the Seattle metropolitan area. These cities are known for their luxury real estate markets, strong school districts, and proximity to major employers like Microsoft and Amazon. Homes in these cities command significantly higher prices, as reflected in their tall and high-positioned boxplots.

In contrast, the bottom 10 cities — including Algona, Pacific, SeaTac, Covington, and Kent — tend to be more affordable residential or working-class areas with lower demand in the housing market. These areas show much lower median prices and narrower price spreads, indicating both limited variability and market value.

This visual evidence confirms the hypothesis that location, particularly proximity to high-demand urban centers and affluent neighborhoods, strongly influences property value

# 3.EXPANDING OUR INVESTMENT KNOWLEDGE 

Beyond the current dataset on individual property listings, broader market-level data can offer critical insights into real estate investment decisions. One highly valuable source is the [Zillow Economic Research Data](https://www.zillow.com/research/data/), which provides aggregated housing market statistics across cities, counties, and states in the U.S. These datasets include metrics such as the Zillow Home Value Index (ZHVI), median rental prices (ZORI), housing affordability, price-to-rent ratios, inventory levels, and median days on market. These indicators are essential for identifying regional price trends, rental potential, and market volatility — all of which are key components of a sound investment strategy.

While our current dataset captures the physical and transactional attributes of properties — such as bedrooms, lot size, renovation status, and sale prices — it lacks macroeconomic context. By integrating Zillow data, we can enrich this analysis with temporal and regional trends. For example, combining property-level prices with Zillow’s home value index over time would help investors identify appreciation patterns and cyclical markets. Additionally, metrics like price-to-rent ratios can help distinguish markets where it is more profitable to buy versus rent, particularly useful for investors considering rental income as part of their ROI.

Incorporating such external datasets allows investors to move beyond one-dimensional decisions based on property features alone. It supports a more holistic view that considers regional growth trends, affordability shifts, and market saturation levels. In essence, pairing transactional-level data with broader market indicators creates a stronger foundation for forecasting returns, reducing investment risk, and identifying emerging opportunities in both high-growth and undervalued areas.

